Cover letter for the original IJRR submission

### 2017-04-20 body/contact.tex
cosmetics
reviewer 2
cosmetics + performance

\subsection{Contact planning algorithm}
Starting from an initial whole-body configuration, we compute a sequence
of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
%~ Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
%~ the contacts are created, allowing it to break the combinatorial.

@@

\subsubsection{Maintaining a contact in the sequence:}

@@

%\subsection{Generation of a contact configuration}  
\subsubsection{Creating contacts:}
Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest. If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others.
%~ \begin{enumerate}
%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
%~ \end{enumerate}

@@

To generate contacts, we proceed as follows.
The contact generator treats each limb $R^k$ independently.
Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
\begin{equation}
\label{eq:contact_gen}
	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \wedge \mathbf{q}^k \in  C_{Contact}^k 
\end{equation}

>>>>

\subsection{Contact planning algorithm}
Starting from an initial whole-body configuration, we compute a sequence
of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
The algorithm can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.

@@

\subsubsection{Maintaining a contact in the sequence}

@@

%\subsection{Generation of a contact configuration}  
\subsubsection{Creating contacts}
Contacts are created using a FIFO approach: we try first to create a contact with the limb that has been contact-free the longest. If the contact creation does not succeeds, the limb is pushed on top of the queue, and will only be tried again after the others.
%~ \begin{enumerate}
%~ \item \deladp{At most one contact creation happens between two consecutive steps. }
%~ \item \deladp{A contact is validated if and only if the resulting configuration is in static equilibrium;}
%~ \item \deladp{We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
%~ was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.}
%~ \end{enumerate}

@@

To generate contacts, we proceed as follows.
The contact generator treats each limb $R^k$ independently.
Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
\begin{equation}
\label{eq:contact_gen}
	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \textrm{ \textbf{and}}\ \mathbf{q}^k \in  C_{Contact}^k 
\end{equation}

### 2017-04-20 body/overview.tex
cosmetics + performance

\subsection{Computation of a guide path --- \Pa}
We first consider the problem of planning a root guide path (Figure~\ref{fig:framework}--$\mathcal{P}_1$).  The dimension of the path is equal to the number of degrees of freedom (DoFs) of the root of the robot.
Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: there must exist a joint configuration that results in static equilibrium for each root configuration\footnote{Enforcing static equilibrium is a classical, conservative approach to reduce the search space and considering states  with non-zero accelerations and velocities, which can't be connected trivially. This does not mean that the final motion will necessary be quasi-static.}. Previous work verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration. We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.

An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot to allow contact creation; not too close, because the robot must avoid collision.
%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision (Figure~\ref{fig:reach_cond} - red shape), while the reachable workspace is in collision (Figure~\ref{fig:reach_cond} - green shapes).
%~ Verifying the \textit{reachability condition} is very efficient in practice.
To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
This is detailed in Section~\ref{rbprm}.
%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).

@@

In the remainder of this paper, we use the terms \gls{contact reachable} and \gls{equilibrium feasible} to qualify either a root configuration, a whole-body configuration, or a set
of such configurations.

\subsection{Generating a discrete sequence of contact configurations --- \Pb}
The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
Each root configuration is then extended into a whole-body configuration in static equilibrium.
The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.
It takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually. 
%~ Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
%~ Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.

The details of the contact sequence generation are given in Section~\ref{sec:contact}.

@@

\medskip
\textbf{A robot} is a kinematic tree $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
Therefore $R$ is fully described by a configuration $\mathbf{q} \in  \mathbb{R}^{r+n}$, with $n$ the number of joint DoFs.
$\mathbf{q}$ is decomposed as follows:
\begin{itemize}
	\item $\mathbf{q}^k$ is a configuration (a vector of joint values) of the limb $R^k$; % (Figure~\ref{fig:character});
	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$. We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ is the world coordinates vector of $R^0$.
\end{itemize}
%Finally, $ \mathbf{q}_{start}$ and  $\mathbf{q}_{goal}$ are the start and goal configurations given as an input of our problem.

%\subsubsection{Environment and other volumes} 
%\label{sec:A}
\medskip
\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}), or equal to it. $W^0$ is typically a low-polygonal bounding shape
of $R^0$ for performance reasons.\\
 %~ (Figure~\ref{fig:character}-right: central cylinder). 
 $W^k$ is the reachable workspace of a limb $R^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
\begin{equation}
  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{j.lim}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
\end{equation}
%~ \adnote{in this definition you don't account for joint limits}
where $\mathbf{p}^k$ denotes the end-effector position (in the root frame) of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}.

Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
$W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.

@@

%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
\begin{figure}
  \centering
  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
  \caption{
           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
		   \label{fig:HyQ_roms}
\end{figure}

>>>>

\subsection{Computation of a guide path --- \Pa (Section~\ref{rbprm})}
We first consider the problem of planning a root guide path (Figure~\ref{fig:framework}--$\mathcal{P}_1$).  The dimension of the path is equal to the number of degrees of freedom (DoFs) of the root of the robot.
Similarly to previous work~\citep{Bouyarmane2009} the path must be \textit{equilibrium feasible}: there must exist a joint configuration that results in static equilibrium for each root configuration\footnote{Enforcing static equilibrium is a classical, conservative approach to reduce the search space and considering states  with non-zero accelerations and velocities, which can't be connected trivially. This does not mean that the final motion will necessary be quasi-static.}. Previous works verifiy \textit{equilibrium feasibility} by explicitly computing such a configuration. We preserve the low dimensionality of the problem by approximating \textit{equilibrium feasibility} with \textit{contact rechability}, illustrated in the following.

An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot to allow contact creation; not too close, because the robot must avoid collision.
%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision (Figure~\ref{fig:reach_cond} - red shape), while the reachable workspace is in collision (Figure~\ref{fig:reach_cond} - green shapes).
%~ Verifying the \textit{reachability condition} is very efficient in practice.
To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).

@@

In the remainder of this paper, we use the terms \gls{contact reachable} and \gls{equilibrium feasible} to qualify either a root or a whole-body configuration, or a set
of such configurations.

\subsection{Generating a discrete sequence of contact configurations --- \Pb (Section~\ref{sec:contact})}
The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
Each root configuration is then extended into a whole-body configuration in static equilibrium.
The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.
It takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually. 
%~ Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
%~ Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.

%~ The details of the contact sequence generation are given in Section~\ref{sec:contact}.

@@

\medskip
\textbf{A robot} is a kinematic tree $R$ composed of: a root $R^0$, and $l$ limbs $R^k, 1 \leq k \leq l$, attached to the root.
The root has $r \geq 6$ DoFs: for instance, HRP-2 has two extra DoFs in the torso, such that we have $r=8$.
Thus $R$ is fully described by a configuration (a vector of joint values) $\mathbf{q} \in  \mathbb{R}^{r+n}$, with $n$ the number of joint DoFs.
$\mathbf{q}$ is decomposed as follows:
\begin{itemize}
	\item $\mathbf{q}^k$ is a configuration  of the limb $R^k$; % (Figure~\ref{fig:character});
	\item $\mathbf{q}^{\overline{k}}$ is a vector of joint values of R \textbf{not} related to $R^k$. We define for convenience \mbox{$\mathbf{q}= \mathbf{q}^k \oplus \mathbf{q}^{\overline{k}}$}; %Abusively, we denote by $\mathbf{q}^0 \oplus \mathbf{q}^k$ any configuration respecting $\mathbf{q}^0$ and $\mathbf{q}^k$;
	\item $\mathbf{q}^{0}\in \mathbb{R}^r$ is the world coordinates vector of $R^0$.
\end{itemize}
%Finally, $ \mathbf{q}_{start}$ and  $\mathbf{q}_{goal}$ are the start and goal configurations given as an input of our problem.

%\subsubsection{Environment and other volumes} 
%\label{sec:A}
\medskip
We then define a set of 3d volumes  $W^i, 0 \leq i \leq l$, each  attached to one joint of the root, such that 
$W^i(\mathbf{q}^{0})$ describes the world position of $W^i$ for the root configuration $\mathbf{q}^{0}$.

\noindent $W^0$ is a volume encompassing $R^0$ (Figure~\ref{fig:HyQ_roms}), or equal to it\footnote{$W^0$ is typically a low-polygonal bounding shape
of $R^0$ for performance.}.\\
 %~ (Figure~\ref{fig:character}-right: central cylinder). 
 $W^k$ is the reachable workspace of a limb $R^k$: %(Figure~\ref{fig:character}-right: the four ellipses). 
\begin{equation}
  W^k = \left\{ {\mathbf{x} \in \mathbb{R}^3: \exists \, \mathbf{q}^k \in C^k_{j.lim}, \mathbf{p}^k(\mathbf{q}^k) = \mathbf{x} } \right\}
\end{equation}
%~ \adnote{in this definition you don't account for joint limits}
where $\mathbf{p}^k$ denotes the end-effector position (in the root frame) of $R^k$ (translation only) for $\mathbf{q}^0 = \mathbf{0}$ being the null displacement, and  $C^k_{j.lim}$ is the space
of admissible limb joint configurations. We also define \mbox{$W = \bigcup_{k=1}^{l}W^k$}.

%~ Each $W^i, 0 \leq i \leq l$ is  attached to one joint of the root, such that 
%~ $W^i(\mathbf{q}^{0})$ is the volume $W^i$ transformed the by root configuration $\mathbf{q}^{0}$.

@@

%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
\begin{figure}
  \centering
  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
  \caption{
           Reachable workspace and torso bounding box of HyQ. Each green shape represent a reachable workspace $W^k$ of a limb. The red shape is $W^0$.}
		   \label{fig:HyQ_roms}
\end{figure}

### 2017-04-20 body/results.tex
cosmetics + performance

At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$ for these scenarios.

@@

\subsection{Description of the scenarios}
In all the scenarios considered, the formulation of the problem is always the same:
a start and goal root configurations are provided as input.
The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
In each scenario we detail the contacts involved and the heuristics chosen (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, all of which are defined in the Appendix~\ref{sec:heuristics}).
 %~ and the constraints on the reachable workspaces (for instance in all the scenarios, the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
%~ The companion video presents the complete contact sequence obtained in all these scenarios.

% \subsubsection*{Truck egress (Figure~\ref{res_truck_pres} and Figure~\ref{res_truck_bd}) -- Humanoid and insectoid robots.}
%\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:stair_robust}):}

@@

\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing}):}
From a bent configuration, the robot has to stand up using a wall as support, and climbing a 25-cm high step.

@@

\subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car}):}
In this scenario inspired from the DRC car egress HRP-2 has to step out of the Polaris car.

@@

\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs}):}
In this long scene, HyQ has to cross a 55-cm large hole, followed by a narrow ``bridge'', only 25-cm large.

@@

\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning}):}
In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.

@@

\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot}):}
This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.

@@

\subsection{Performance analysis} \label{sec:perf}
To analyze performance, we ran the planner 1000 times for each considered scenario.
We measured the computation time spent in each part of the algorithm, and analyzed their success rate.

@@

\subsubsection{Computation times:}
Table~\ref{tab:requestime} summarizes the computation times.

@@

\subsubsection{Success rates:}
Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
Despite the complexity of the scenarios addressed, as well as the approximations made in our formulation, our planner succeeded in the large majority of cases.

@@

\subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
To validate our plans, we either use the framework proposed in \cite{Carpentier2016} or our own implementation of a $\mathcal{P}_3$ solver, detailed in Appendix~\ref{app:optim}.
The companion video shows the obtained motions.

@@

>>>>

At the end of the video, we validate our contact plans by exhibiting a solution to the interpolation problem $\mathcal{P}_3$.

@@

\subsection{Description of the scenarios}
In all the scenarios considered, the formulation of the problem is always the same:
a start and goal root configurations are provided as input (except in the stair climbing scenario where the start whole body configuration is given).
The framework computes the initial contact configuration, and outputs a sequence of contact configurations connecting it to the goal.
In each scenario we detail the contacts involved and the heuristics chosen (either $h_{\textrm{\it EFORT}}$, $h_{vel}$ or $h_{w}$, all of which are defined in the Appendix~\ref{sec:heuristics}).
 %~ and the constraints on the reachable workspaces (for instance in all the scenarios, the reachable workspaces of the legs of HRP-2 are always required to intersect with the environment). 
%~ The companion video presents the complete contact sequence obtained in all these scenarios.

% \subsubsection*{Truck egress (Figure~\ref{res_truck_pres} and Figure~\ref{res_truck_bd}) -- Humanoid and insectoid robots.}
%\subsubsection*{Truck egress -- Humanoid and insectoid robots (Figure~\ref{res_truck_bd}).}
\subsubsection{HRP-2 -- Steep staircase (Figure~\ref{fig:stair_robust})}

@@

\subsubsection{HRP-2 -- Standing up (Figure~\ref{fig:standing})}
From a bent configuration, the robot has to stand up using a wall as support, and climbing a 25-cm high step.

@@

\subsubsection{HRP-2 -- Polaris car egress (Figure~\ref{fig:car})}
In this scenario inspired from the DRC car egress HRP-2 has to step out of the Polaris car.

@@

\subsubsection{HyQ -- Obstacle race (Figure~\ref{fig:HyQ_bridge} and~\ref{fig:HyQ_obs})}
In this long scene, HyQ has to cross a 55-cm large hole, followed by a narrow ``bridge'', only 25-cm large.

@@

\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning})}
In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.

@@

\subsubsection{3-fingered hand -- Manipulation of a pen (Figure~\ref{fig:penrot})}
This scenario is proposed to illustrate the generality of our approach: we consider a manipulation task for a robotic hand and use
our contact planner to compute a contact sequence for the fingers, considered as effectors (Figure~\ref{fig:penrot}).
Although we do not address the hard issue of accounting for rolling motions, the planner is able to compute the shown sequences in less than 5 seconds.

@@

\subsection{Performance analysis} \label{sec:perf}
To analyze performance, we ran the planner 1000 times for each scenario.
We measured the computation time spent in each part of the algorithm, and analyzed their success rate.

@@

\subsubsection{Computation times (Table~\ref{tab:requestime})}
 %~ summarizes the computation times.

@@

\subsubsection{Success rates (Table~\ref{tab:sucess_planning})}
%~ Table~\ref{tab:sucess_planning} summarizes the successful planning rates.
Despite the complexity of the scenarios and the approximations made in our formulation, our planner succeeded in the large majority of cases.

@@

\subsection{Validation of the contact plans, and comparison with previous work} \label{sec:compa}
To validate our plans, we either use the framework of \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
The companion video shows the obtained motions.

@@

To provide a completely fair comparison, it would be required to compare the methods on the same computers and exact same scenarios, which is not possible
because the source code of the other methods is often not available. Despite this observation and the fact that some results are old, we believe that the results show that our method
performs better. Indeed, a difference in technology does not justify a difference as important as the one demonstrated in Table~\ref{tab:sucess_planning}.
